Simulating High-Dimensional Multivariate Data

### 2020-11-04 050-examples.Rmd
revise order of code chunks/text so have dynamic inline R
draft quad loss example

This section demonstrates how to simulate multivariate data using `bigsimr`, aiming to replicate the structure of high-dimensional dependent count data.
Simulating RNA-sequencing (RNA-seq) data is a primary motivating application of the proposed methodology, seeking scaleable Monte Carlo (MC) methods for realistic multivariate simulation for these data.
Intergene correlation is an inherent part of biological processes.
Yet many models do not account for this, leading to major distruptions to the operating characteristics of statistical estimation, testing, and prediction.
See @Efron2012 for a detailed discussion with related methods and see @Wu2012b, @Schissler2018, @Schissler2019 for applied examples.
The following subsections apply `bigsimr`'s methods to real RNA-seq data, including replicated an estimated parameteric structure, MC probability estimation, and MC correlation estimation efficiency assessment.

@@

```{r ch050-readBRCA, echo=FALSE, eval=TRUE}
## full workflow simulating RNA-seq data
brca <- readRDS(file = brcaRDS)
## smaller set for prototyping
## d=5
## brca <- brca[ , 1:d]
```

In an illustration of our proposed methodology applied to real data, we seek to simulate RNA-sequencing data by producing simulated random vectors with assumed marginal distributions with estimated parameters.
Our goal is to replicate the structure of a breast cancer data set (BRCA data set from The Cancer Genome Atlas).
Specifically, we will simulate $B=10,000$ random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$
All these genes exhibit over-dispersion and, so, we proceed to estimate the NB parameters $(r_i, p_i), i=1,\ldots,d$ to determine the target marginal PMFs $g_i(y_i)$ (via method of moments).
Often researchers posit a negative binomial (NB) model as RNA-seq counts are often over-dispersed that a Poisson model would suggest.
This suggests simulating high-dimensional multivariate NB (MVB) with heterogeneous marginals would be useful tool in the development and evaluation of RNA-seq analytics.
This procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions. 
Notably, the marginal NB probabilities $\hat{p}_i's$ are small --- ranging in $[`r min(probs)` , `r max(probs)`]$.
This gives rise to highly variable counts and, typically, less restriction on potential pairwise correlation pairs.
To complete the simulation algorithm inputs, we estimate the Spearman correlation matrix $\bf{R^{Spearman}_Y}$ and set that as the target correlation.

@@

```{r ch050-estRhoBRCA, echo=TRUE, eval=TRUE}
## 1. Estimate Spearman's correlation on the count data
## corType <- 'pearson'
corType <- 'spearman'
system.time( nb_Rho <- bigsimr::cor_fast( brca, method = corType ) )
```

@@

Once the functions are defined/executed to complete marginal estimation, we specify targets and generate the desired random vectors using `rvec`:

@@

```{r ch050-runBRCA, echo=FALSE, results=FALSE, cache = FALSE}
brcaSimRDS <- paste0( "results/brca", round(myProb*100), "sim.rds" )
if( !file.exists(brcaSimRDS) ) {
    ## Set the number of random vectors
    n <- 10000
    ## construct margins
    nb_margins <- make_nbinom_alist(sizes, probs)
    ## run sims
    system.time( sim_nbinom <- rvec(n, nb_Rho, nb_margins, type = corType,
                                    ensure_PSD = TRUE, cores = cores) )
    colnames(sim_nbinom) <- names(brca)
    saveRDS( sim_nbinom, brcaSimRDS )
}
sim_nbinom <- readRDS( brcaSimRDS )
```

@@

>>>>

This section demonstrates how to simulate multivariate data using `bigsimr`, aiming to replicate the structure of high-dimensional dependent count data.
Simulating RNA-sequencing (RNA-seq) data is a primary motivating application of the proposed methodology, seeking scaleable Monte Carlo (MC) methods for realistic multivariate simulation for these data.
Intergene correlation is an inherent part of biological processes.
Yet many models do not account for this, leading to major distruptions to the operating characteristics of statistical estimation, testing, and prediction.
See @Efron2012 for a detailed discussion with related methods and see @Wu2012b, @Schissler2018, @Schissler2019 for applied examples.
The following subsections apply `bigsimr`'s methods to real RNA-seq data, including replicated an estimated parameteric structure, MC probability estimation, and MC evaluation of correlation estimation efficiency.

@@

```{r ch050-readBRCA, echo=FALSE, eval=TRUE}
## full workflow simulating RNA-seq data
## brcaRDS <- 'data/brca99.rds'
brca <- readRDS(file = brcaRDS)
## smaller set for prototyping
## d=5
## brca <- brca[ , 1:d]
```

In an illustration of our proposed methodology applied to real data, we seek to simulate RNA-sequencing data by producing simulated random vectors with assumed marginal distributions with estimated parameters.
Our goal is to replicate the structure of a breast cancer data set (BRCA data set from The Cancer Genome Atlas).
Specifically, we will simulate $B=10,000$ random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$
All these genes exhibit over-dispersion and, so, we proceed to estimate the NB parameters $(r_i, p_i), i=1,\ldots,d$ to determine the target marginal PMFs $g_i(y_i)$ (via method of moments).
Often researchers posit a negative binomial (NB) model as RNA-seq counts are often over-dispersed that a Poisson model would suggest.
This suggests simulating high-dimensional multivariate NB (MVB) with heterogeneous marginals would be useful tool in the development and evaluation of RNA-seq analytics.
This procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions. 
To specify the simulation algorithm inputs, we estimate the Spearman correlation matrix $\bf{R^{Spearman}_Y}$ and marginal NB parameters.

@@

```{r ch050-estRhoBRCA, echo=TRUE, eval=TRUE}
## Estimate Spearman's correlation on the count data
corType <- 'spearman'
system.time( nb_Rho <- bigsimr::cor_fast( brca, method = corType ) )
```

@@

Notably, the marginal NB probabilities $\hat{p}_i's$ are small --- ranging in $[`r min(probs)` , `r max(probs)`]$.
This gives rise to highly variable counts and, typically, less restriction on potential pairwise correlation pairs.
Once the functions are defined/executed to complete marginal estimation, we specify targets and generate the desired random vectors using `rvec`:

@@

```{r ch050-runBRCA, echo=FALSE, results=FALSE, cache = FALSE}
nb_margins <- make_nbinom_alist(sizes, probs)
brcaSimRDS <- paste0( "results/brca", round(myProb*100), "sim.rds" )
if( !file.exists(brcaSimRDS) ) {
    ## Set the number of random vectors
    n <- 10000
    ## construct margins
    ## run sims
    system.time( sim_nbinom <- rvec(n, nb_Rho, nb_margins, type = corType,
                                    ensure_PSD = TRUE, cores = cores) )
    colnames(sim_nbinom) <- names(brca)
    saveRDS( sim_nbinom, brcaSimRDS )
}
sim_nbinom <- readRDS( brcaSimRDS )
```

@@

## MC evaluation of correlation estimation efficiency

```{r ch010-myM, include=FALSE, cache = TRUE}
m <- 100
```

MC methods are routinely used in many statistical inferential tasks including estimation, hypothesis testing, error rates, and empirical interval coverage rates.
For an concise introduction to these methods, see, for example @Rizzo2007, Ch. 6.
Now, we demonstrate how `bigsimr` can be used evaluate estimation efficiency.
In particular, we'd like to assess the error our correlation estimation above.
We used a conventional method, based on classical statistical theory. 
Yet this method was not designed for high-dimensional data.
Indeed, high-dimensional covariance estimation (and precision matrices) is an active area of statistical science (see, for example, @Won2013g and @VanWieringen2016).
In this example, we simulate $m=`r m`$ data sets with the number of simulated vectors matching the number of patients in the BRCA data set, $N=`r nrow(brca)`$.
At each iteration we estimate the quadratic loss from the specified $\bf{R^{Spearman}_Y}$ and average over the $m$ iteration.

```{r ch050-quadloss, echo=TRUE, eval=TRUE, cache=FALSE}
quadLossRDS <- paste0( "results/brca", round(myProb*100), "quadLoss.rds" )
if ( !file.exists( quadLossRDS) )  {
    n <- nrow(brca)
    simRho <- replicate(n = m, expr = { tmpSim <- rvec(n = nrow(brca) , nb_Rho, nb_margins,
                                                       type = corType, ensure_PSD = TRUE, cores = cores)
                                                       tmpRho <- bigsimr::cor_fast( x = tmpSim )} ,
                        simplify = FALSE)

    quadLoss <- unlist( lapply( simRho, rags2ridges::loss, T = nb_Rho, type = "quadratic"  ) )
    saveRDS(quadLoss, quadLossRDS)
}
quadLoss <- readRDS( quadLossRDS )
```

The standard `R` summary function supplies the mean-augmented five-number summary of the quadratic loss distribution computed above.

```{r ch050-quadloss-summary, echo=TRUE, eval=TRUE, cache=FALSE}
summary(quadLoss)
```

This distribution could be compared to more elaborate, high-dimensional estimators to guide analysts in deciding whether the additional complexity and computation time are warranted for their application.
